---
title: "Predator Prey Body Size Prior Prediction"
author: "Justin Pomeranz - J Pomz"
date: "11/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Predator-Prey Body Sizes

Understanding predator-prey interactions has long been a research interest of ecologists. Body size is related to a number of aspects which influences these interactions. For example, predators are often though to be gape-limited, meaning that larger predators should be able to consume larger prey. 

For this example, let's examine the assumption that the mean prey body mass increases log-linearly with predator body mass. In other words, let's assume that larger predators consume larger prey. We will be examining a simple linear model in the form of $$y_i = a + b * x_i$$ 

where $y_i$ is the mean prey mass, $x_i$ is the mean predator mass, and $a$ and $b$ are the intercept and slope variables, repsectively.

In this linear model, we will assume that the distribution of prey mass and both of the intercept and slope variables folow a normal distribution. For the sigma variable, we will assume it follows an exponential distribution. 

The full Bayesian model can be described as:
$$  M_{prey} \sim  dnorm(\mu, \sigma) $$
$$    \mu = \alpha + \beta * M_{pred} $$
$$    \alpha \sim dnorm(0, 5) $$
$$    \beta \sim dnorm(0, 5) $$
$$    \sigma \sim dexp(1) $$

# Body size data

The data set of Brose et al. (2006) documents over 10,000 predator-prey interactions, including the mean mass of each. This data set will be used to demonstrate the concept of prior predictive simulations. For this example, we will be using the `rethinking` package. 

```{r, message = FALSE}
library(rethinking)
```

First, let's load the data:

```{r}
d <- readRDS("~/Wesner_Post_Doc/prior_prediction/data/pred-prey-mass.RDS")
```

This data table has been cleaned and only includes rows where mean mass for both the predator and prey was recorded. 

To simplify this example, we will be working with log-transformed variables:

```{r}
d$log_pred <- log(d$pred)
d$log_prey <- log(d$prey)
```

## Prior Information

Before looking at the data, let's consider what we *expect* to see. First, let's consider the intercept. Remember that we are working with data on the log scale. With that in mind, the intercept in this model tells us what the average prey mass is for a predator with mass of 0 on the log scale, which equates to a mass of 1: 

```{r}
exp(0)
``` 

For a starting point, let's assume that predators are essentially the same size as their prey. That would mean that the intercept variable, \alpha, would be centered near 0 on the log-log scale. We will return to the intercept variable for a consideration of the standard deviation of the distribution. 

Now that we have an idea of where the intercept should be centered, let's consider the relationship between prey mass and predator mass. Continuing with the basic assumption that prey are approximately the same size as their predators, an expected slope value of 1 seems reasonable. In other words, the average prey body size increases exactly directly in response to predator mass. 

# Flat Priors
First, let's consider flat priors (sometimes called vague priors). Flat priors are commonly used in the literature, and are often justified because they have little influence on the posterior distributions. Priors are "flattened" by giving them a large standard deviation. An example of a flat prior for a normal distribution would be Normal(0, 10) or even Normal(0,100).

For both the intercept variable, \alpha, and the slope variable, \beta, let's consider a flat prior with a standard deviation of 10. Therefore, our prior distributions are Normal(0, 10) and Normal(1, 10)

In order to assess these priors, let's simulate some data 

```{r}
# set.seed for reproducibilty
set.seed(1055)
N <- 100 # number of samples
a <- rnorm(N, 0, 10) # prior distribution of alpha
b <- rnorm(N, 1, 10) # prior distribution of beta

# set plot axis limits
xlim <- c(-25, 25)
ylim = c(-25, 25)
# setup the plot
plot(NULL, xlim = xlim, ylim = ylim,
     xlab = "log predator mass", ylab = "log prey mass")
mtext(("a ~ rnorm(0,10)      b ~ rnorm(1,10)"))

# plot 100 lines based on the prior distributions
for (i in 1:N){
  curve(a[i] + b[i]*x,
        from = -25,
        to = 25,
        add = TRUE,
        col = gray(0, 0.3))
}
```
Remember that this is on a log-log scale. In order to put the overall range into perspective, let's add reference lines for the weight of a virus and a blue whale. On the log scale this is approximately -20.7 and 18.8, respectively.

```{r, echo = FALSE}
# setup the plot
plot(NULL, xlim = xlim, ylim = ylim,
     xlab = "log predator mass", ylab = "log prey mass")
mtext(("a ~ Normal(0,10)      b ~ Normal(1,10)"))

# plot 100 lines based on the prior distributions
for (i in 1:N){
  curve(a[i] + b[i]*x,
        from = min(xlim),
        to = max(xlim),
        add = TRUE,
        col = gray(0, 0.3))
}
# virus
abline(a = -20.7, b = 0, col = "blue", lwd = 2)
abline(v = -20.7, col = "blue", lwd = 2)
# blue whale
abline(a = 18.8, b = 0, col = "blue", lwd = 2, lty = 2)
abline(v = 18.8, col = "blue", lwd = 2, lty = 2)
```
Now, the solid lines indicate the weight of a typical virus, and the dashed lines represent the weight of a blue whale. 

For reference, let's add a 1:1 line to the plot. The 1:1 line is what we would expect to see if the mean prey mass was exactly equal to the predator mass. 

```{r, echo = FALSE}
# setup the plot
plot(NULL, xlim = xlim, ylim = ylim,
     xlab = "log predator mass", ylab = "log prey mass")
mtext(("a ~ Normal(0,10)      b ~ Normal(1,10)"))

# plot 100 lines based on the prior distributions
for (i in 1:N){
  curve(a[i] + b[i]*x,
        from = min(xlim),
        to = max(xlim),
        add = TRUE,
        col = gray(0, 0.3))
}
# 1:1 line
abline(a = 0, b = 1, col = "red", lty = 2)
# virus
abline(a = -20.7, b = 0, col = "blue", lwd = 2)
abline(v = -20.7, col = "blue", lwd = 2)
# blue whale
abline(a = 18.8, b = 0, col = "blue", lwd = 2, lty = 2)
abline(v = 18.8, col = "blue", lwd = 2, lty = 2)
```
Consider what these simulations represent. A considerable amount of the lines are impossibly steep. For example, as predator size increases ~10 log units, the priors consider it reasonable that the prey size will vary by nearly 40 log units (-20 to 20). Even though we centered our slopes at 1, having a flat SD of 10 means that 95% of the Gaussian probability mass lies between -19 and 19. We can surely do better than this. Let's maintain our assumption that mean prey size approximately increases in a 1:1 relationship, but we will bound it with an SD of 1. This puts 95% of the probability mass between -1 and 3. These are still quite vague, and allow for some incredibly strong relationships. 

Now, let's consider the intercept value. we used a prior with a normal distribution with a mean of 0 and an SD of 10. Let's plot this to visualize what this implies. 

```{r}
plot(density(a))
```
With the large SD value, the model is considering intercepts from -20 to 20 as reasonable. An intercept of -20 essentially means that the mean prey size for a 1 gram predator is approximately the size of a virus. Likewise, an intercept of +20 means that the mean prey size for the same 1 g predator is larger than a blue whale. We can surely limit this to do better, but first, let's reconsider the mean value. 

Centering the intercept at 0 seemed like a reasonable first approximation. However, it is probably more realistic that predators will have to be slightly larger than their prey, meaning that our distribution should place more probability mass below 0. Without looking at the data, it's hard to have a strong prior estimation of how much smaller prey will be than their predators. Let's assume that predators are on average one logarithmic unit smaller than their prey, so we will center the intercept distribution at -1. For the Standard Deviation, let's use a smaller, but still somewhat vague prior of 2.5. This will put 95% of the probability mass between -6 and 4.  

# Informative priors

```{r, echo = FALSE}
plot_prior <- function(N = 100,
                       a_mean,
                       a_sd, 
                       b_mean, 
                       b_sd,
                       xlim,
                       ylim) {
  set.seed(1055)
  a <- rnorm(N, a_mean, a_sd)
  b <- rnorm(N, b_mean, b_sd)
  
  plot(NULL, xlim = xlim, ylim = ylim,
       xlab = "log predator mass", ylab = "log prey mass")
  mtext(paste0("a ~ Normal(", a_mean, ",", a_sd, ")",
               "  b ~ Normal(", b_mean, ",", b_sd, ")"))
  abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
  for (i in 1:N){
    curve(a[i] + b[i]*x,
          from = min(xlim),
          to = max(xlim),
          add = TRUE,
          col = gray(0, 0.3))
  }
}

plot_prior(a_mean = -1,
           a_sd = 2.5,
           b_mean = 1,
           b_sd = 1,
           xlim = xlim,
           ylim = ylim
           )
# virus
abline(a = -20.7, b = 0, col = "blue", lwd = 2)
abline(v = -20.7, col = "blue", lwd = 2)
# blue whale
abline(a = 18.8, b = 0, col = "blue", lwd = 2, lty = 2)
abline(v = 18.8, col = "blue", lwd = 2, lty = 2)
```
This is certainly looking more reasonable. Most of the intercepts are below the 1:1 line, but there are still quite a few above it, meaning that the model will be testing a range of values, and the priors won't be overwhelming the data. 

However, I would posit that the range of slopes is still impossibly large. Let's try one more time by shrinking the SD value to 0.5, placing 95% of the probability mass between 0 and 2. 

```{r, echo = FALSE}
plot_prior(a_mean = -1,
           a_sd = 2,
           b_mean = 1,
           b_sd = 0.5,
           xlim = xlim,
           ylim = ylim)
# virus
abline(a = -20.7, b = 0, col = "blue", lwd = 2)
abline(v = -20.7, col = "blue", lwd = 2)
# blue whale
abline(a = 18.8, b = 0, col = "blue", lwd = 2, lty = 2)
abline(v = 18.8, col = "blue", lwd = 2, lty = 2)

```

Now let's fit a model, just to illustrate what we actually find. 

```{r}
m1 <- quap(
  alist(
    log_prey ~ dnorm(mu, sigma),
    mu <- a + b * log_pred,
    a ~ dnorm(-1, 2),
    b ~ dnorm(1, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)
precis(m1)

post <- extract.samples(m1, 20)

plot(d$log_pred,
     d$log_prey,
     xlim = range(d$log_pred),
     ylim = range(d$log_prey),
     xlab = "log predator mass",
     ylab = "log prey mass",
     col = rangi2)
mtext("a ~ Normal(-1, 2)",
              "  b ~ Normal(1, 0.5")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
title("Posterior Regression lines N = 20")

for (i in 1:20){
  curve(post$a[i] + post$b[i]*x,
        # from = min(xlim),
        # to = max(xlim),
        add = TRUE,
        col = gray(0, 0.3))
}
```

